A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Swap S5.1 and S5.2

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
\Cref{sec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
This indicates that for a given raw term and context, any two synthesised types must be equal.
This property will also be used in the proof of the logical decidability of bidirectional type synthesis.
Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.

\subsection{Mode Correctness}\label{sec:mode-correctness}
As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every typing rule:
\begin{enumerate*}
\item each `input' type variable in a premise must be synthesised from `earlier' premises or provided by the conclusion, if the rule is checking;
\item each `output' type variable in the conclusion should have been synthesised from premises already, if the rule is synthesis.
\end{enumerate*}
Here an `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
It is important to note that the order of premises in a bidirectional typing rule matters, as it also sets the order of bidirectional type synthesis.

@@

As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property \emph{cannot} be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
Now that we have the notion of signature available, we can articulate mode-correctness more precisely.

But, first, we assume that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
For clarity in presentation, however, we will continue to use named variables.

Now we proceed with the definition of mode-correctness.
As we have outlined at the beginning, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Hence, we instead choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The finite subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined inductively by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align*}
    \synvar(\cdot)                                  & = \emptyset  \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align*}
\end{definition}
This subset accumulates type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness}
  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
  \begin{align*}
    \MC_{\mathit{as}}( \cdot ) ={} & \top \\
    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}) 
    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
    \land \fv(\Delta) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \right)\\
    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk})
    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})\right).
  \end{align*}
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an 'output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of 'input' variables in its conclusion.
This condition must be met for every tail of the argument list to ensure that `output' variables accessible at each argument position are from preceding arguments only.
\begin{definition}
  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
Finally, for a checking operation, an 'input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each instance of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.

\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every raw term $\isTerm{t}$ in synthesising mode and two derivations of 
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
\end{lemma}

\begin{theorem}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode decide whether
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
  \]
  has a derivation for some type $A$.
\end{theorem}

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every context $\Gamma$ and raw term $\isTerm{t}$, exactly one of the following statements holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.

@@

>>>>

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
We start with presenting the statement of the decidability of bidirectional synthesis (whose proof will be deferred until \Cref{subsec:bidirectional-synthesis-checking}) and the trichotomy on raw terms in \Cref{subsec:trichotomy}.
We will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to exhibit the trichotomy as a corollary.
We will then formalise the mode-correctness condition and show the uniqueness of synthesized types in \Cref{sec:mode-correctness}.
This uniqueness property means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
This will be used later in the proof of the decidability of bidirectional type synthesis and checking in \Cref{subsec:bidirectional-synthesis-checking}.

\begin{theorem}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  it is decidable whether there exists some $A$ such that
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
  \]
\end{theorem}

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  Exactly one of the following statements holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.

\subsection{Mode Correctness}\label{sec:mode-correctness}
As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every typing rule:
\begin{enumerate*}
\item each `input' type variable in a premise must be synthesised from `earlier' premises or provided by the conclusion, if the rule is checking;
\item each `output' type variable in the conclusion should have been synthesised from premises already, if the rule is synthesis.
\end{enumerate*}
Here an `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised types are accumulated during type synthesis.

@@

Now we proceed with the definition of mode-correctness.
As we have outlined at the beginning, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Hence, we instead choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align*}
    \synvar(\cdot)                                  & = \emptyset  \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align*}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness}
  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
  \begin{align*}
    \MC_{\mathit{as}}( \cdot ) ={} & \top \\
    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right).
  \end{align*}
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}
  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
Finally, for a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.

\begin{remark}
  As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property cannot be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
  Now that we have the notion of signature available, we can articulate mode-correctness above.
\end{remark}

\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}

@@

We assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
For clarity in presentation, however, we will continue to use named variables.
